Treebeard : An Optimizing Compiler for Decision Tree Based ML Inference

### 2022-04-25 docs/MICRO 2022/overview.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\section{Treebeard Overview}
Figure \ref{Fig:CompilerStructure} shows the high level structure of \Treebeard{}. 
The input to \Treebeard{} is a serialized decision tree ensemble. 
%% Popular frameworks like XGBoost and LightGBM are supported and the system is easily extensible to other frameworks.
Given an input model our compiler generates optimized inference code. Specifically it generates a 
callable batch inference function \texttt{predictForest} that, given an array of input rows, computes an
array of model predictions.

\Treebeard{} specializes the code generated for inference by progressively optimizing and lowering a 
high level representation of the \texttt{predictForest} function down to LLVM IR \cite{LLVM}.
%% By \textbf{\emph{lowering}}, we mean 
\textbf{\emph{Lowering}} is the process of transforming an operation at a higher 
abstraction to a sequence of operations at a lower abstraction. Optimizations in \Treebeard{} 
are implemented using a combination of annotation and lowering. An operation at a higher 
abstraction is annotated with attributes that indicate what kind of optimization is 
to be performed while lowering it. The lowering transformation uses this information to generate 
optimized lower level code. For example, tree tiling and loop ordering (figure \ref{Fig:Overview} 
shows two possible loop orders -- one tree at a time and one row at a time) are decided 
at the highest abstraction. These decisions are communicated to the lowering pass 
which explicitly encodes them in the lowered IR as shown in Figure \ref{Fig:Overview}.

%Treebeard first constructs a high level representation of the decision forest inference operation.
\Treebeard{}'s IR has three abstraction levels as shown in Figure \ref{Fig:Overview}.  
At the highest level (HIR), the input model is represented as a collection of binary trees. This is shown 
in the second row of Figure \ref{Fig:Overview}. At this level of abstraction,
\Treebeard{} tiles nodes together to transform a binary tree into an $n$-ary tree 
and decides what order trees are to be traversed in. In the example in the figure, 
trees are tiled with a tile size of 2. Tiling is indicated using colored elipses drawn 
around nodes that are in the same tile. Tree reordering is another transformation that is performed at this abstraction. 
One objective of tree reordering is to group identically structured trees so that they can share the same traversal code
to improve locality in the instruction cache.  Alternative objectives for reordering can easily be supported.
With the tiling shown in Figure \ref{Fig:Overview}, Tree1 and Tree3 have depth 1, while Tree2 has a depth of 2.
Therefore, the trees are reordered so that Tree1 and Tree3 are together.

The high level IR is then lowered to a mid-level IR (MIR). The aim of MIR is 
to allow optimizations that are independent of the final memory layout of the model. 
The lowering to MIR performs loop transformations like loop tiling, permutation, parallelization and unrolling.
At this level of the IR (the third row in Figure \ref{Fig:Overview}),
the order in which trees, input row pairs are walked is explicitly represented in the IR using loop nests.
Figure \ref{Fig:Overview} shows two possible ways that loop nests could be generated -- the first
walks all rows for one tree before moving to the next tree while the second walks all trees for 
one row before moving to the next row. Another aspect handled by this lowering is the fissing of 
loops to make sure that trees with the same structure share code. Here, both versions of MIR shown 
ensure that Tree1 and Tree3 share the same traversal code, while traversal code for Tree2 is different.
Additionally, tree walk optimizations, such as tree walk unrolling (shown in figure \ref{Fig:Overview}) and 
peeling are performed on the MIR.

\Treebeard{} then further lowers the IR to explicitly represent the memory layout of the model. Buffers to hold model 
values are inserted into the generated code and all tree operations in the mid-level IR are lowered to explicitly 
reference these buffers. Additionally, at this level of the IR, \Treebeard{} generates vectorized code to take 
advantage of SIMD instructions where applicable. Finally, the inference function is translated to LLVM IR and 
JIT compiled to executable code.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/compiler.pdf}
  \caption{Treebeard compiler structure}
  \label{Fig:CompilerStructure}
\end{figure}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{figures/OverviewExample.pdf}
  \caption{\Treebeard{} IR lowering and optimization details. The three abstraction levels in Treebeard's IR are shown. The
           high level IR is a tree based IR to perform model level optimization, the mid-level IR is for
           loop optimizations that are independent of memory layout and the low level IR allows us to perform
           vectorization and other memory layout dependent optimizations.}
  \label{Fig:Overview}
\end{figure*}

@@

% \subsection{Mid Level IR}
% The Mid Level IR optimizes the loop structures and tree walks. Firstly, the order in which the iteration space of trees and inputs is walked is determined and by re-ordering the loop nesting if necessary. Also, operations such as \texttt{isLeaf}, \texttt{traverseTile}, \texttt{getLeafValue} are introduced so that the traversal of trees explicitly represented. The following listing shows the IR for inference using a model with four trees on an input batch with two rows. The listed IR walks all trees for one input row before moving to the next row. One important point to note here is that details such as the data structure used for the trees are not explicitly encoded in the IR. This allows us to reuse optimization and lowering passes on this level of the IR regardless of what the final in memory representation of the model is.

@@

% \subsection{Low Level IR}
% The IR is finally lowered to a form where the in memory representation of the model is made explicit. Buffers to hold the model values are inserted into the generated code and all tree operations in the mid-level IR are lowered to explicitly reference these buffers. The semantics of all operations are made explicit. For example, \texttt{traverseTreeTile} is lowered into a series of operations to load thresholds, feature indices and features, compare the features with the thresholds and compute the next node to evaluate. This IR is then lowered directly to LLVM IR and JITted.

>>>>

\section{Treebeard Overview}
\label{Sec:Overview}
Figure \ref{Fig:CompilerStructure} shows the high level structure of \Treebeard{}. 
The input to \Treebeard{} is a serialized decision tree ensemble. 
%% Popular frameworks like XGBoost and LightGBM are supported and the system is easily extensible to other frameworks.
Given an input model our compiler generates optimized inference code. Specifically, it generates a 
batch inference function, \texttt{predictForest} that, given an array of input rows, computes an
array of model predictions.

\Treebeard{} specializes the code generated for inference by progressively optimizing and lowering a 
high level representation of the \texttt{predictForest} function down to LLVM IR \cite{LLVM}.
%% By \textbf{\emph{lowering}}, we mean 
\textbf{\emph{Lowering}} is the process of transforming an operation at a higher 
abstraction to a sequence of operations at a lower abstraction. Optimizations in \Treebeard{} 
are implemented using a combination of annotation and lowering. An operation at a higher 
abstraction is annotated with attributes that guide how code should be lowered. 
Actual transformation happens while lowering.
%The lowering transformation uses this information to generate optimized lower level code.
 For example, tree tiling and loop ordering (Figure \ref{Fig:Overview} 
shows two possible loop orders -- one tree at a time and one row at a time) are decided 
at the highest abstraction. These decisions are communicated to the lowering pass 
which explicitly encodes them in the lowered IR as shown in the Figure.

%Treebeard first constructs a high level representation of the decision forest inference operation.
\Treebeard{}'s IR has three abstraction levels as shown in Figure \ref{Fig:CompilerStructure}.  
At the highest level (HIR), the input model is represented as a collection of binary trees. This is shown 
in the second row of Figure \ref{Fig:Overview}. At this level of abstraction,
\Treebeard{} tiles nodes together to transform a binary tree into an $n$-ary tree. 
In the example in the Figure, 
trees are tiled with a tile size of 2, as indicated using colored ellipses drawn 
around nodes that are in the same tile.

Tree reordering is another transformation that is performed at this abstraction. 
One objective of tree reordering is to group identically structured trees so that they can share the same traversal code
to improve locality in the instruction cache.  Alternative objectives for reordering can easily be supported.
With the tiling shown in Figure \ref{Fig:Overview}, Tree1 and Tree3 have depth 1, while Tree2 has a depth of 2.
Therefore, the trees are reordered so that Tree1 and Tree3 are together.

The high level IR is then lowered to a mid-level IR (MIR).
The lowering to MIR also performs loop transformations like loop tiling, permutation, parallelization and unrolling.
At this level of the IR (the third row in Figure \ref{Fig:Overview}),
the order in which trees, input row pairs are walked is explicitly represented in the IR using loop nests.
Figure \ref{Fig:Overview} shows two possible ways that loop nests could be generated.
\CommentOut{
 -- the first
walks all rows for one tree before moving to the next tree while the second walks all trees for 
one row before moving to the next row.
}
 Another aspect handled by this lowering is the fissing of 
loops to make sure that trees with the same structure share code. Here, both versions of MIR shown 
ensure that Tree1 and Tree3 share the same traversal code, while traversal code for Tree2 is different.

 The aim of MIR is 
to allow optimizations that are independent of the final memory layout of the model. 
Tree walk optimizations, such as tree walk unrolling (shown in Figure \ref{Fig:Overview}) and 
peeling are performed on the MIR.

\Treebeard{} then further lowers the IR to explicitly represent the memory layout of the model. Buffers to hold model 
values are inserted into the generated code and all tree operations in the mid-level IR are lowered to explicitly 
reference these buffers. Additionally, at this level of the IR, \Treebeard{} generates vectorized code. 
Figure~\ref{Fig:Overview} shows operations for which SIMD instructions are eventually generated. 
Finally, the inference function is translated to LLVM IR and 
JIT compiled to executable code.

We implement \Treebeard{} using the MLIR compiler infrastructure\cite{MLIR}. MLIR provides modular and reusable infrastructure
to develop intermediate representations for domain-specific compilers. Firstly, to enable compiler writers to 
quickly design and build intermediate representations, MLIR has a large set of 
\textbf{\emph{dialects}} that provide operations from various domains. Each dialect 
contains a logically related set of operations and types; dialect transformations and conversions allow lowering and conversion of these operations and types to other lower-level dialects and subsequently translation to LLVM IR. 
For example, the \texttt{scf} dialect provides control flow constructs like loops and if-else operations. 
The \texttt{memref} dialect provides functionality to manage buffers. We build \Treebeard{} using a custom
MLIR dialect and a combination of dialects provided in MLIR. Secondly, MLIR provides an extremely powerful 
operator rewriting infrastructure. We make extensive use of this to implement transformations.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/compiler}
  \caption{\Treebeard{} compiler structure.}
  \label{Fig:CompilerStructure}
\end{figure}

\begin{figure*}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/OverviewExample_New.pdf}
  \vskip 10pt
  \caption{\Treebeard{} IR lowering and optimization details: the three abstraction levels in \Treebeard{}'s IR are shown. The
           high level IR is a tree-based IR to perform model level optimization, the mid-level IR is for
           loop optimizations that are independent of memory layout and the low level IR allows us to perform
           vectorization and other memory layout dependent optimizations.}
  \label{Fig:Overview}
\end{figure*}

@@

% \subsection{Mid Level IR}
% The Mid Level IR optimizes the loop structures and tree walks. Firstly, the order in which the iteration space of trees and inputs is walked is determined and by re-ordering the loop nesting if necessary. Also, operations such as \texttt{isLeaf}, \texttt{traverseTile}, \texttt{getLeafValue} are introduced so that the traversal of trees explicitly represented. The following listing shows the IR for inference using a model with four trees on an input batch with two rows. The listed IR walks all trees for one input row before moving to the next row. One important point to note here is that details such as the data structure used for the trees are not explicitly encoded in the IR. This allows us to reuse optimization and lowering passes on this level of the IR regardless of what the final in-memory representation of the model is.

@@

% \subsection{Low Level IR}
% The IR is finally lowered to a form where the in-memory representation of the model is made explicit. Buffers to hold the model values are inserted into the generated code and all tree operations in the mid-level IR are lowered to explicitly reference these buffers. The semantics of all operations are made explicit. For example, \texttt{traverseTreeTile} is lowered into a series of operations to load thresholds, feature indices and features, compare the features with the thresholds and compute the next node to evaluate. This IR is then lowered directly to LLVM IR and JITted.

### 2022-04-25 docs/MICRO 2022/probtiling.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\subsection{Probability Based Tiling}
\label{sec:ProbTiling}
%\subsubsection*{Motivation}

The next algorithm we propose exploits the inherent biases among the leaves of a decision tree. In typical machine learning models some leafs (equivalently outcomes or predictions) are more likely to be reached than others. In such settings, having balanced tiled trees is not sufficient to minimize expected inference time.

@@

On the other hand, for the benchmark \op{epsilon},
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the training input.
This means that most trees in \op{epsilon} are not leaf biased.

\CommentOut{
\subsubsection{More Notation}
In order to formulate the probability based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}
}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \caption{Statistical profile for airline-ohe}
    \label{Fig:AirlineOHEStats}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \caption{Statistical profile for epsilon}
    \label{Fig:EpsilonStats}
\end{figure}
\CommentOut{
We say that an input row $r_i$ is \textbf{\emph{covered}} by a subset of leaves $L' \subseteq L$ of a tree $T$, if the leaf $l$ reached by 
walking $T$ for row $r_i$ is in $L'$. We show how different models (and even different trees within the same model) behave differently 
using models for two benchmarks, airline-ohe and epsilon. Consider the graphs shown in   
figures \ref{Fig:AirlineOHEStats} and \ref{Fig:EpsilonStats}. Each line in these graphs corresponds to a fixed fraction of the input (say $f$). 
A point on a line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a fraction $f$ of all training inputs with a fraction $x$ of 
leaves. For example, the first point on the $f=0.9$ line in figure \ref{Fig:AirlineOHEStats} says that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure \ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction of inputs for the benchmark airline-ohe. 
This means that a small fraction of leaves are very likely. On the other hand, for the benchmark epsilon,
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the test input.
This means that most trees in the epsilon model are not leaf biased.
One other observation we make is that most models have some leaf biased trees while the rest of the trees have equally likely leaves.
\TODO{AP Maybe define a term for trees with roughly equally likely leaves?} We design the probability based tiling algorithm to take advantage of this property 
of decision tree ensembles. 
}
\subsubsection{The Optimization Problem}

@@

Observe that the latency of one tree walk is proportional to the number of tiles that need to be evaluated to reach the leaf. It is easy to see that for a leaf biased tree, basic tilling does not optimize for this objective, it considers all leafs to be equally likely.

@@

The above optimization problem can be solved optimally using dynamic programming. 
We leave this out in the interest of space. 
Instead, we use the simple greedy algorithm listed in algorithm \ref{Alg:GreedyTilingAlgo} to construct a valid tiling given the node probabilities\footnote{Probabilites for internal nodes can be computed from probablities for leafs by summing up the probabilities of all leafs that belong to the sub-tree rooted at the internal node. Leaf probabilities are collected during training.}.
The algorithm starts at the root and greedily keeps adding the most probable legal node to the current tile until the maximum tile size is reached.
Subsequently, the tiling procedure is recursively performed on all nodes that are destinations for edges going out of the constructed tile.

% \subsubsection{Dynamic Programming Formulation}

@@

% Intuitively, it seems like the following greedy algorithm also gives the optimal tiling. The algorithm starts at the root and greedily keeps adding the most probable node to the current tile until the maximum tile size is reached.
\begin{algorithm}
    \caption{Greedy Probability Based Tree Tiling}
    \label{Alg:GreedyTilingAlgo}
    \begin{algorithmic}
        \Procedure{TileTree}{$\Tree = (V, E, r)$, $n_t$} 
            \If {$r \in L_{\Tree}$}
                \State \textbf{return} $\{ r \}$
            \EndIf
            \State $Tile \leftarrow \{ r \}$
            \While{$|Tile| < n_t$}
                \State $e = (u,v) \in Out(Tile)$ st $p(v)$ is max and $v \notin L$
                \If{$e = \emptyset$}
                    \State \textbf{break}
                \EndIf
                \State $Tile = Tile \cup \{ v \}$
            \EndWhile
            \State $Tiles =  \{ Tile \}$
            \For{$(u,v) \in Out(Tile)$}
                \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
            \EndFor
            \State \textbf{return} $Tiles$
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

% Talk about problems with increasing number of tile shapes and only performing such tiling on skewed trees
\CommentOut{
When we tried to apply algorithm \ref{Alg:GreedyTilingAlgo} on all trees in our benchmarks, we found
that even minor variations in probability caused the tiling algorithm to generate a large 
number of tile shapes. This in turn caused a loss in performance because the large size of the 
lookup table needed (section \ref{sec:LookupTable}) caused increased L1 cache misses. In order to 
alleviate this, we only perform probability based tiling on trees that are leaf biased.
}
We find probability based tiling is only beneficial for leaf biased trees\footnote{Turns out that for trees that are not leaf biased, probability based tiling produces many more tile shapes (see Section~\ref{sec:tileShapes}) which direclty impacts the cost of \op{getChildTile} making it more expensive than basic tiling.}.  Recall that   
a tree to be leaf biased if a small fraction of leaves, say $\alpha$, can cover a large fraction of training inputs, say $\beta$.
We only perform probability based tiling on trees with thresholds $\alpha=0.05$ and $\beta=0.9$ and fall back to uniform tiling otherwise.

>>>>

\subsection{Probability-Based Tiling}
\label{sec:ProbTiling}
%\subsubsection*{Motivation}

\CommentOut{
%% We have already given the motivation for probabilistic tiling!
The next algorithm we propose exploits the inherent biases among the leaves of a decision tree. In typical machine learning models some leaves (equivalently outcomes or predictions) are more likely to be reached than others. In such settings, having balanced tiled trees is not sufficient to minimize expected inference time.

@@

On the other hand, for the benchmark \op{epsilon},
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the training input.
This means that most trees in \op{epsilon} are not leaf biased.
}

\CommentOut{
\subsubsection{More Notation}
In order to formulate the probability-based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}
}

\CommentOut{
\begin{figure}
  \subfloat[airline-ohe]{
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \label{Fig:AirlineOHEStats}
  }
  \subfloat[epsilon]{
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \label{Fig:EpsilonStats}
  }
  \caption{Statistical profiles.}
\end{figure}
}

\CommentOut{
We say that an input row $r_i$ is \textbf{\emph{covered}} by a subset of leaves $L' \subseteq L$ of a tree $T$, if the leaf $l$ reached by 
walking $T$ for row $r_i$ is in $L'$. We show how different models (and even different trees within the same model) behave differently 
using models for two benchmarks, airline-ohe and epsilon. Consider the graphs shown in   
figures \ref{Fig:AirlineOHEStats} and \ref{Fig:EpsilonStats}. Each line in these graphs corresponds to a fixed fraction of the input (say $f$). 
A point on a line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a fraction $f$ of all training inputs with a fraction $x$ of 
leaves. For example, the first point on the $f=0.9$ line in figure \ref{Fig:AirlineOHEStats} says that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure \ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction of inputs for the benchmark airline-ohe. 
This means that a small fraction of leaves are very likely. On the other hand, for the benchmark epsilon,
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the test input.
This means that most trees in the epsilon model are not leaf biased.
One other observation we make is that most models have some leaf biased trees while the rest of the trees have equally likely leaves.
\TODO{AP Maybe define a term for trees with roughly equally likely leaves?} We design the probability-based tiling algorithm to take advantage of this property 
of decision tree ensembles. 
}

\CommentOut{
\subsubsection{The Optimization Problem}

@@

Observe that the latency of one tree walk is proportional to the number of tiles that need to be evaluated to reach the leaf. It is easy to see that for a leaf biased tree, basic tilling does not optimize for this objective, it considers all leaves to be equally likely.

@@

The above optimization problem can be solved optimally using dynamic programming. 
We leave this out in the interest of space. 
Instead, we use the simple greedy algorithm listed in algorithm \ref{Alg:GreedyTilingAlgo} to construct a valid tiling given the node probabilities\footnote{Probabilites for internal nodes can be computed from probablities for leaves by summing up the probabilities of all leaves that belong to the sub-tree rooted at the internal node. Leaf probabilities are collected during training.}.
The algorithm starts at the root and greedily keeps adding the most probable legal node to the current tile until the maximum tile size is reached.
Subsequently, the tiling procedure is recursively performed on all nodes that are destinations for edges going out of the constructed tile.
}
% \subsubsection{Dynamic Programming Formulation}

@@

% Intuitively, it seems like the following greedy algorithm also gives the optimal tiling. The algorithm starts at the root and greedily keeps adding the most probable node to the current tile until the maximum tile size is reached.
\begin{algorithm}
    \caption{Greedy probability-based tree tiling.}
    \label{Alg:GreedyTilingAlgo}
    \small{
    \begin{algorithmic}[1]
        \Procedure{TileTree}{$\Tree = (V, E, r)$, $n_t$} 
            \If {$r \in L_{\Tree}$}
                \State \textbf{return} $\{ r \}$
            \EndIf
            \State $Tile \leftarrow \{ r \}$
            \While{$|Tile| < n_t$}
                \State $e = (u,v) \in Out(Tile)$ st $p(v)$ is max and $v \notin L$
                \If{$e = \emptyset$}
                    \State \textbf{break}
                \EndIf
                \State $Tile = Tile \cup \{ v \}$
            \EndWhile
            \State $Tiles =  \{ Tile \}$
            \For{$(u,v) \in Out(Tile)$}
                \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
            \EndFor
            \State \textbf{return} $Tiles$
        \EndProcedure
    \end{algorithmic}
    }
\end{algorithm}

The goal of probability-based tiling is to minimize the average inference latency, or equivalently to minimize the expected
number of tiles that are evaluated to compute one tree prediction
We present a simple greedy heuristic algorithm (Algorithm~\ref{Alg:GreedyTilingAlgo}) to construct a valid tiling given
the node probabilities\footnote{Probabilities for internal nodes can be computed from probabilities for leaves by summing
up the probabilities of all leaves that belong to the sub-tree rooted at the internal node. Leaf probabilities are collected
during training.}. The algorithm starts at the root and greedily keeps adding the most probable legal node to the current
tile until the maximum tile size is reached. Subsequently, the tiling procedure is recursively performed on all nodes 
that are destinations for edges going out of the constructed tile. 
In the algorithm, these edges are denoted by $Out(Tile)$ and the subtree rooted at a node $v$ is denoted by $S_v$.
It is easy to see that the tiling constructed by Algorithm \ref{Alg:GreedyTilingAlgo} is valid.

% Talk about problems with increasing number of tile shapes and only performing such tiling on skewed trees
\CommentOut{
When we tried to apply algorithm \ref{Alg:GreedyTilingAlgo} on all trees in our benchmarks, we found
that even minor variations in probability caused the tiling algorithm to generate a large 
number of tile shapes. This in turn caused a loss in performance because the large size of the 
lookup table needed (section \ref{sec:LookupTable}) caused increased L1 cache misses. In order to 
alleviate this, we only perform probability-based tiling on trees that are leaf biased.
}

As  probability-based tiling is only beneficial for leaf-biased trees,
%\footnote{Turns out that for trees that are not leaf biased, probability-based tiling produces many more tile shapes (see Section~\ref{sec:tileShapes}) which direclty impacts the cost of \op{getChildTile} making it more expensive than basic tiling.}.
%% Recall that a tree to be leaf biased if a small fraction of leaves, say $\alpha$, can cover a large fraction of training inputs, say $\beta$.
we  perform probability-based tiling on trees only when a small fraction ($\alpha$) of leaves cover a large part ($\beta$) of the training inputs.
% In our experiments we set $\alpha=0.075$ and $\beta=0.9$ and fall back to basic tiling otherwise.
